---
title: "Generating Summary Tables"
author: "Eric Moore"
date: "Last Update: `r Sys.Date()`"

output: 
  html_document:
      theme: simplex
      highlight: kate
      df_print: paged
---

This guide provides instructions on how researchers can create standardized descriptive tables for their projects. It is recommended that researchers utilize RStudio to run the various code chunks to obtain the summary tables, though there are alternative applications that can be utilized. If you do not have R and Rstudio already installed on your computer, please utilize the following links for instructions for your respective operating system:

* [Windows](https://www.geeksforgeeks.org/how-to-install-r-studio-on-windows-and-linux/)
* [Mac OS](https://www.makeuseof.com/install-r-mac/)

It is also recommended that researchers make as many changes to their dataset as possible in the statistical software package of their choice and then export said dataset as a Comma-delimited (csv) file. If applicable, researchers should ensure that value labels are saved for requisite variables instead of data values. The process for doing this is outlined for the three most common statistical programs below:

* [SPSS](https://www.ibm.com/docs/en/spss-statistics/beta?topic=files-exporting-datasets)
* [STATA](https://www.stata.com/manuals/dimportdelimited.pdf#dimportdelimited)
* [SAS](https://documentation.sas.com/doc/en/egug/8.3/n0umuuda0dou3jn1lvmyq6uv834h.htm)

After ensuring RStudio is properly installed and the dataset has been edited or changed as necessary, it is now time to begin setting up the environment to run the summary tables.

## Setting the Environment

This document provides instructions for researchers on how to use the code in the SuTables_Code Markdown file that can be found in the Github Folder. Prior to running the code to generate the tables, however, we first need to set up the environment. This includes setting a few options for knitr (RMarkdown's tool to generate code), loading the requisite packages, and creating the theme for the tables.

```{r setup, warning=F, message=F}
# DO NOT CHANGE THESE OPTIONS
knitr::opts_chunk$set(echo = TRUE)
options(knitr.duplicate.label = "allow")

rm(list = ls())

# Libraries
library(tidyverse)
library(haven)
library(readr)
library(modelsummary)
library(gt)
library(kableExtra)
library(webshot2)
library(pdftools)

# Theme for Tables, based on the apa theme created for GT
# source: https://gist.github.com/pdparker/1b61b6d36d09cb295bf286a931990159
apa_style <- function(data) {
  data %>%
    opt_table_lines(extent = "none") %>%
    tab_options(
      table.align = "left",
      heading.border.bottom.width = 2,
      heading.border.bottom.color = "black",
      heading.border.bottom.style = "solid",
      table.border.top.color = "white",
      table_body.hlines.color = "white",
      table_body.border.top.color = "black",
      table_body.border.top.style = "solid",
      table_body.border.top.width = 1,
      heading.title.font.size = 16,
      table.font.size = 12,
      heading.subtitle.font.size = 14,
      heading.align = "left",
      table_body.border.bottom.color = "black",
      table_body.border.bottom.width = 2,
      table_body.border.bottom.style = "solid",
      column_labels.border.bottom.color = "black",
      column_labels.border.bottom.style = "solid",
      column_labels.border.bottom.width = 1,
      data_row.padding = px(5)
    ) %>%
      opt_table_font(font = "georgia")
}
```

```{r table-functions, warning=F, message=F}
# Paste these functions into your R document (DO NOT ALTER)

# Code for creating summary table on all numeric variables in the dataset
Allnumeric2 <- function(df){
datasummary(All(df, numeric=TRUE) ~ 
            ((Obs. = N) + (Avg. = Mean) + (Std.Dev. = SD) + (Min. = Min) + (Max. = Max)),
            data = df,
            output = 'gt') %>% 
  apa_style() %>% 
  cols_align(align = "center", columns = 2:(ncol(df) - 1))  %>% 
  tab_header(title = "Table Number", subtitle = md("*Change to a Descriptive Title for the Table*") ) %>% 
  tab_footnote(footnote = md( "*Note*: Possible footnote for table.") ) %>%
  tab_footnote(footnote = "Second possible footnote for table.")  %>%
  tab_options(table.width = pct(50))
}

# Code for creating  summary table for all categorical variables in the dataset
Allcat2 <- function(df){
datasummary(All(df, numeric=FALSE, factor=TRUE) ~ 
            ((Obs. = N) + Percent() ),
            data = df,
            output = 'gt') %>% 
  apa_style() %>% 
  cols_align(align = "center", columns = 3:4)  %>% 
  tab_header(title = "Table Number", subtitle = md("*Change to a Descriptive Title for the Table*") ) %>%   
  tab_footnote(footnote = md( "*Note*: Possible footnote for table.") ) %>%
  tab_footnote(footnote = "Second possible footnote for table.")  %>%
    cols_width(
    columns = c(1) ~ px(100),
    columns = c(2:4) ~ px(75)  ) 
}

# Code for creating  summary table for all numeric variables by a group variable
Crossnum <- function(df, var1){
datasummary(All(df, numeric=TRUE) ~ 
            var1 * ((Avg. = Mean) + (Std.Dev. = SD)),
            data = df,
            output = 'gt') %>% 
  apa_style() %>% 
  cols_align(align = "center", columns = 2:(ncol(df) - 2))  %>% 
  tab_header(title = "Table Number", subtitle = md("*Change to a Descriptive Title for the Table*") ) %>% 
  tab_footnote(footnote = md( "*Note*: Possible footnote for table.") ) %>%
  tab_footnote(footnote = "Second possible footnote for table.")  %>%
  tab_options(table.width = pct(50))
}

# Code for creating  summary table for all categorical variables by a group variable
Crosscat <- function(df, var1){
datasummary(All(df[!myvars], numeric=FALSE, factor=TRUE, character = FALSE) ~ 
            var1 * ((Obs. = N) + Percent() ) ,
            data = df,
            output = 'gt') %>% 
  apa_style() %>% 
  cols_align(align = "center", columns = 3:(ncol(df) - 2))  %>% 
  tab_header(title = "Table Number", subtitle = md("*Change to a Descriptive Title for the Table*") ) %>% 
  tab_footnote(footnote = md( "*Note*: Possible footnote for table.") ) %>%
  tab_footnote(footnote = "Second possible footnote for table.")  %>%
  tab_options(table.width = pct(50))
}

# Code for creating combined summary table (numeric and categorical) by a group variable
Combinedtable <- function(df, var1){
var1 <- ensym(var1)  
fac <- as.formula(paste("~", as.character(var1)))

datasummary_balance(fac,
                    data = df, output = 'gt',
                    dinm = F) %>% 
  apa_style() %>%
  cols_align(align = "center", columns = 3:(ncol(df) - 1))  %>% 
  tab_header(title = "Table Number", subtitle = md("*Change to a Descriptive Title for the Table*") ) %>%   
  tab_footnote(footnote = md( "*Note*: Possible footnote for table.") ) %>%
  tab_footnote(footnote = "Second possible footnote for table.") %>%
  tab_options(table.width = pct(60)) #  %>%  gtsave("table_example.pdf")
}
```

## Getting Data into the R Environment

The next step in the process is to load your dataset and see the variables you want to include in the descriptive table. To load data into the R environment using RStudio, complete the following steps:

* From the Environment tab click on the Import Dataset Menu
* Select the file extension from the option
* Either enter the file name or select "browse" in the top line of the prompt.
* Once the file has been selected, name the dataset under the import options box and click "import".

The following represents an example of manually importing a dataset, and performing a few modifications or changes:
```{r data}

# The following code brings in a test dataset in order to run the tables: 
example <- 'https://vincentarelbundock.github.io/Rdatasets/csv/Ecdat/Treatment.csv'
example <- read.csv(example)

example <- example %>%
           mutate(PriorInc. = re75 / 1000,
                  PostInc. = re78 / 1000,
                  Treatment = ifelse(treat == TRUE, 'Treatment', 'Control'),
                  Married = ifelse(married == TRUE, 'Yes', 'No')) %>%
            dplyr::select(PriorInc., PostInc., Treatment,
                   Ethnicity = ethn, Age = age, Education = educ, Married)
```
<br> 
There are some additional steps you may need to take in order to ensure that your data will run properly with the provided code below. These commands are provided below, and can be copied and pasted as needed:
```{r cleaning-data}
# This converts the the dataset into a dataframe
# Necessary to run some of the functions for datasummary 
df2 <- data.frame(rbind(example))

# Convert all Character Variables to Factors
df2[sapply(df2, is.character)] <- lapply(df2[sapply(df2, is.character)],  as.factor)

# Replace 0 or missing value with NA
df2 <- df2 %>% mutate_if(is.numeric, ~replace(., . == 0, NA))

# Replace possible infinite values with NA
df2[sapply(df2, is.infinite)] <- NA

# Also need to set this initial option, where you list the dataset of interest 
#and the variable that you want to utilize as your "grouping" variable
myvars <- names(df2) %in% c("Treatment") 
```

## General Information about Creating Tables

The following chunks of code provide researchers the opportunity to have consistent looking tables when providing summary statistics on both categorical and numerical variables. The functions that are utilized below are provided in the Rmarkdown file under the code section title "table-functions". You will not be able to run the functions in this section until you have copied and pasted the appropriate table function and run it in your R Document. 

One additional thing to note in the proceeding code chunks is the option of gtsave. This option saves the table as a pdf document in your working directory (For information on changing the working directory for your project, reference this webpage from [Postit Support](https://support.posit.co/hc/en-us/articles/200711843-Working-Directories-and-Workspaces-in-the-RStudio-IDE). Delete the # symbol in the final line of code and change the table_example line to a descriptive title. 

## Creating Tables for All Variables

### Numeric Variables

The first line of code represent a function that returns a table on all numeric variables in the dataset. The dataframe (here titled df2) should correspond to the dataframe the researcher has imported into R.

It is important to note that part of the code is currently "hidden" by being commented (gtsave). Uncomment this option by deleting the # sign when you are satisfied that the table meets your expectations.

```{r table:numeric1}
Allnumeric2(df2)  # %>% gtsave("Table1.pdf")
```
### Categorical Variables

Researchers may also want to create a basic table comprised of just their categorical variables. For this we utilize the second function in the code chunk of the RMD file and obtain the following output: 
```{r table:categorical}
Allcat2(df2) # %>% gtsave("Table2.pdf")
```

### Crosstabs
There are occasions when the previous two tables may only provide some of the information that is pertinent for the researcher or the project. In some cases, it may be beneficial to provide a cross tabulation to explore the relationship or distribution between a specific grouping variable, e.g. we may want to see the difference between treatment and control groups.  

This first line of code provides a cross tabulation of all numeric variables in the dataset over a group variable, in this case Treatment. Researchers would simply change df2 (dataset) and Treatment (the group variable) to be applicable to their dataset:
```{r table:numcross}
Crossnum(df2, df2$Treatment) # %>% gtsave("Table3.pdf")
```
This second line of code provides a cross tabulation of all categorical variables in the dataset over a group variable. Once again, researchers will change df2 (dataset) and Treatment (the group variable) as they would in the previous line of code:
```{r table:numcat}
Crosscat(df2, df2$Treatment) # %>% gtsave("Table4.pdf")
```

### Bringing Everything Together

Instead of including multiple tables delineated by the type of variables, however, researchers may want to create one combined table that plots both numeric and categorical variables across one grouping variable. The code below utilizes the final function in the previous code chunk to create the standardized "Table 1" in many social science papers. Note that there are now two options the researcher must select. The first continues to be the dataframe of interest (df2), but the second option is the grouping variable in the dataset (here named Treatment). The code provides the following output:

```{r table:combined}
Combinedtable(df2, Treatment) # %>% gtsave("Table5.pdf")
```

### Merging the files together

Now it is time to bring the tables together as a singular pdf file. Currently, the tables are saved as indiviudal pdf files in the working directory. The following code will combine the specified tables and then create a singular output in the same folder. 

```{r table:pdfoutput, eval=FALSE, include=TRUE}
pdf_combine(c("Table1.pdf", "Table2.pdf", "Table3.pdf"), output = "Tables.pdf")
```
The left side of the code (after pdf_combine(c( ) explicitly lists the tables that you created above once you have uncommented the gtsave option within the code chunks. The right side of the code (after output = ) is the title of the corresponding file that will appear in the working directory folder.  
